---
title: "ESM 244 Lab Week 2"
author: "Danielle Sclafani"
date: "1/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(palmerpenguins)
library(here)

# for PCA
library(ggfortify)

#for ggplot customization
library(readxl)
library(gghighlight)
library(lubridate)
library(plotly)

#to make compound figure
library(patchwork)
```


## Part 1: PCA with penguins - wrangling and scaling the data

```{r}
penguin_pca <- penguins %>% 
  select(body_mass_g, ends_with("_mm")) %>% # select any column that ends in _mm
  drop_na() %>% # to remove in all columns, you do not have to specify within dorp_na
  scale() %>% #all values are scaled
  prcomp() #makes it a PCA

penguin_pca$rotation # run this to see the principle components (PCA)

```

### creating a biplot - using autoplot function

```{r}

penguin_complete <- penguins %>% # we want a data set with all the variables so we can add things like species color to our graph
  drop_na(body_mass_g, ends_with("_mm")) #drop na in columns used in PCA

autoplot(penguin_pca,
         data = penguin_complete,
         colour = "species",
         loadings = TRUE, #adds the arrows
         loadings.label = TRUE) + theme_classic()# adds the arrow labels

#sum of axi variations = total variation predicted by the biplot
```
### Part 2: ggplot customization

##### Read in an excel file and do some wrangling
- graph

```{r}
fish_noaa <- read_excel(here("data","foss_landings.xlsx")) %>% 
  janitor::clean_names() %>% 
  mutate(across(where(is.character), tolower)) %>% #mutate columns across where there is a character to lower case, text in the observations instead of just the names of the column
  mutate(nmfs_name = str_sub(nmfs_name, end = -4)) %>% # remove the last four characters which are **** in the column, do not want to change the name, this will just replace it and not add a new one
  filter(confidentiality == "public")
```

make a customized graph
```{r}
fish_plot <- ggplot(data = fish_noaa, aes(x=year, y = pounds)) +
  geom_line(aes(color = nmfs_name), show.legend = FALSE) +
  theme_minimal()

fish_plot #when a graph has so many lines, it is called a spaghetti plot, but sometimes we want to highlight only a few lines

ggplotly(fish_plot) #makes graph interactive
```

### Use gghighlight to highlight certain series

```{r}
ggplot(data = fish_noaa, aes(x=year, y = pounds, group = nmfs_name)) +
  geom_line() +
  theme_minimal()+
  gghighlight(nmfs_name == "tunas") #highlights the species we are interested in


ggplot(data = fish_noaa, aes(x=year, y = pounds, group = nmfs_name)) +
  geom_line(aes(color = nmfs_name)) +
  theme_minimal()+
  gghighlight(max(pounds)> 1e8) # can highlight based on values- when max value is greater than 1e8, once you add gghihglight, color will only be applied to the ones that you highlight, even if you say color = nmfs_name earlier in the code
```


### Using mutate to transform variables and change the factor order, to order the way we want instead of alphabetically

#### Read in from a URL, use 'lubridate()', `mutate()`, make a graph with months in order. when you download a URL your data will change as the URL is updated. if you want to know which version of the data you used, download the data so it will not change

```{r}
monroe_wt <- read_csv("https://data.bloomington.in.gov/dataset/2c81cfe3-62c2-46ed-8fcf-83c1880301d1/resource/13c8f7aa-af51-4008-80a9-56415c7c931e/download/mwtpdailyelectricitybclear.csv") %>% 
  clean_names()
```

```{r}
monroe_ts <- monroe_wt %>%
  mutate(date = mdy(date)) %>% #makes a new date column, based on the old date column that is formatted as a date 
  mutate(record_month = month(date)) %>%#pull the month from the date column and call it record_month
  mutate(month_name = month.abb[record_month]) %>% #will contain month abbreviation from the number in record_month
  mutate(month_name = fct_reorder(month_name, record_month)) #reorder the months in the order of the record_month


ggplot(data=monroe_ts, aes(x = month_name, y = total_k_wh))+
  geom_jitter()

```

### Part 3L Compounds figure with `patchwork` making compound figures - multiple graphs that are in an order that we choose

```{r}
graph_a <- ggplot(data = penguins, aes(x=body_mass_g, y= flipper_length_mm)) +
  geom_point()

graph_b <- ggplot(data = penguins, aes(x= species, y = flipper_length_mm)) +
  geom_jitter(aes(color = species), show.legend = FALSE)

# use vertical line operator | to put graphs side by side and use / to put graphs one over the other

graph_c <- (graph_a | graph_b) / fish_plot & theme_dark() #the & sign will add change to all three graphs

graph_c

ggsave(here("fig","graph_c_ds.png"), width = 5, height = 6) #saves the last graph that was run
```






